---
title: "Simple gene dropping"
---

Gene dropping follows alleles from a base population through a simulated
pedigree. Each offspring receives one recombined haplotype from its sire and
one from its dam. This provides a small teaching example of inheritance over
multiple generations.

## Base haplotypes

`simulate_gene_drop()` represents a diploid base population with two matrices,
`hap1` and `hap2`. Rows are individuals, columns are markers, and values are
0/1 alleles.

```{r}
library(qgsim)

base_haplotypes <- list(
  hap1 = matrix(
    c(0, 0, 1, 1,
      0, 1, 0, 1,
      1, 0, 1, 0,
      1, 1, 0, 0),
    nrow = 4,
    byrow = TRUE
  ),
  hap2 = matrix(
    c(1, 1, 0, 0,
      1, 0, 1, 0,
      0, 1, 0, 1,
      0, 0, 1, 1),
    nrow = 4,
    byrow = TRUE
  )
)

map <- c(0, 0.2, 0.7, 1)
```

The map gives marker positions in Morgans on one chromosome. This first
version models only a single chromosome.

## Simulate and sample

The base population is generation 0. Selected sires and dams from each
generation produce the next generation. Crossovers are drawn from a Poisson
distribution based on chromosome length, and crossover locations determine
which parental haplotype is transmitted along the chromosome.

```{r}
result <- simulate_gene_drop(
  base_haplotypes = base_haplotypes,
  map = map,
  generations = 3,
  population_size = 4,
  selected_parents = 4,
  offspring_per_mating = 2,
  sample_size = 2,
  seed = 123,
  keep = "sample"
)

result
result$sample
result$haplotypes$hap1
```

The pedigree always contains the base population and all simulated
generations. With `keep = "sample"`, the haplotype matrices contain only the
sampled final-generation individuals. With `keep = "all"`, they contain every
individual in pedigree order. The `haplotype_row` column maps sampled IDs to
the returned haplotype matrices.

## Scope and future implementations

This is a deliberately simple, single-chromosome R implementation for
teaching. It does not represent a complete breeding simulator. A later
backend could replace the internal `drop_gamete_r()` loop with compiled C++ or
Fortran while preserving the user-facing result structure.
